% ============================================================
% Report: Flow-Matching Speech Enhancement via
%         Dynamic Multi-Layer Tokenizer Alignment
% ============================================================
\documentclass[11pt,a4paper]{article}

% ---- Packages ----
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=2.4cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage[numbers]{natbib}

\hypersetup{colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black}

\title{\textbf{Flow-Matching Speech Enhancement via\\Dynamic Multi-Layer Tokenizer Alignment}}
\author{
  Yiheng Chen \and André Dal Bosco\\[3pt]
  \small CSC\_52002\_EP --- IA Générative Multimodale (2025--2026)\\
  \small École Polytechnique\\
  \small Teachers: Vicky Kalogeiton, Xi Wang
}
\date{}

\begin{document}
\maketitle

% ============================================================
\begin{abstract}
We present a flow-matching speech enhancement system that operates in the continuous latent space of two pre-trained audio tokenizers.
A Rectified Flow model learns a direct velocity field from noisy DAC latents to clean DAC latents, conditioned on features extracted from the MOSS-Audio-Tokenizer.
Our core contribution is a \emph{multi-layer conditioning} strategy: instead of using only the last hidden layer of the MOSS tokenizer (standard practice), we extract all 32 hidden-layer outputs from the deepest encoder stage and fuse them via a learnable weighted sum.
We further propose a \emph{time-dependent fusion} variant where a lightweight MLP maps the ODE timestep to per-layer weights, allowing different denoising stages to emphasise different feature types.
A controlled ablation across four conditioning strategies shows consistent, monotonic improvements in PESQ, STOI, and FAD, confirming that multi-layer tokenizer features provide complementary information for speech enhancement.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

Speech enhancement---the task of extracting clean speech from a noisy recording---is fundamental to many downstream applications including automatic speech recognition, hearing aids, and voice communication.
Classical approaches operate in the time-frequency domain with supervised regression~\cite{lu2013speech}, while recent work has explored generative models that can produce natural-sounding speech by modelling the distribution of clean signals~\cite{richter2023speech}.

A promising recent direction is latent-space generative modelling: rather than operating on raw waveforms or spectrograms, one encodes audio into a compact continuous latent space using a pre-trained codec and performs generation there.
This dramatically reduces the sequence length and allows the generative model to focus on high-level structure.

In this project, we adopt \textbf{Rectified Flow}~\cite{lipman2023flow}---a flow-matching framework that learns a straight-line velocity field between two distributions---to map noisy audio latents to clean audio latents.
A \textbf{Diffusion Transformer} (DiT)~\cite{peebles2023scalable} serves as the velocity predictor, with conditioning injected via cross-attention from features of the \textbf{MOSS-Audio-Tokenizer}~\cite{ma2024language}, a large-scale (1.6B parameter) hierarchical audio encoder.

\paragraph{Core contribution.}
Pre-trained feature extractors contain rich information distributed across their layers: early layers capture fine-grained acoustic details while deeper layers encode more abstract semantic and prosodic features.
In standard practice, only the \emph{last} hidden layer is used for conditioning.
We hypothesise that leveraging \emph{all} layers can improve enhancement quality.
We design two fusion strategies---static and time-dependent---and conduct a systematic four-way ablation to isolate the effect of multi-layer conditioning.

% ============================================================
\section{Background}
\label{sec:background}

\subsection{Rectified Flow}
\label{sec:rectified-flow}

Rectified Flow~\cite{lipman2023flow,liu2023flow} defines a generative process by learning a velocity field that transports samples along straight paths between a source distribution and a target distribution.
Given paired samples $\mathbf{x}_0 \sim p_{\text{source}}$ and $\mathbf{x}_1 \sim p_{\text{target}}$, the interpolated point at time $t \in [0,1]$ is:
\begin{equation}
  \mathbf{x}_t = t\,\mathbf{x}_1 + (1 - t)\,\mathbf{x}_0
  \label{eq:interpolation}
\end{equation}
The target velocity field is constant along each path:
\begin{equation}
  \mathbf{v}^*(\mathbf{x}_t, t) = \mathbf{x}_1 - \mathbf{x}_0
  \label{eq:target-v}
\end{equation}
A neural network $\mathbf{v}_\theta$ is trained to predict this velocity via the MSE objective:
\begin{equation}
  \mathcal{L} = \mathbb{E}_{t \sim U(0,1)}\left[\left\|\mathbf{v}_\theta(\mathbf{x}_t, t, \mathbf{c}) - (\mathbf{x}_1 - \mathbf{x}_0)\right\|^2\right]
  \label{eq:loss}
\end{equation}
where $\mathbf{c}$ denotes optional conditioning information.
At inference, $\mathbf{x}_0$ (noisy latent) is integrated forward using the Euler method with $N$ steps to produce the predicted clean latent $\hat{\mathbf{x}}_1$.

\textbf{Crucially}, in our setting $\mathbf{x}_0$ is \emph{not} Gaussian noise but directly the noisy DAC latent.
The flow learns a denoising transformation in latent space, starting from a meaningful corrupted signal rather than pure noise.

\subsection{Audio Tokenizers}
\label{sec:tokenizers}

\paragraph{DAC} (Descript Audio Codec)~\cite{kumar2024high} compresses 16\,kHz audio to a continuous latent representation at 50\,Hz with dimensionality $d=1024$, followed by residual vector quantisation (RVQ).
We use the \emph{pre-quantisation} continuous latent as both the source ($\mathbf{x}_0$, from noisy audio) and target ($\mathbf{x}_1$, from clean audio) for the flow model.

\paragraph{MOSS-Audio-Tokenizer}~\cite{ma2024language} is a large hierarchical audio encoder (1.6B parameters) organised into 4 stages of progressively increasing abstraction:
\begin{itemize}[itemsep=2pt]
  \item \textbf{Stage 0}: Patch embedding (factor 240) $\to$ 12-layer Transformer, $d=768$
  \item \textbf{Stage 1}: Patch (factor 2) $\to$ 12-layer Transformer, $d=768$
  \item \textbf{Stage 2}: Patch (factor 2) $\to$ 12-layer Transformer, $d=768$
  \item \textbf{Stage 3}: Patch (factor 2) $\to$ \textbf{32-layer Transformer}, $d=1280$
\end{itemize}
The total temporal downsampling is $240 \times 2 \times 2 \times 2 = 1920$, yielding 12.5\,Hz features from 24\,kHz input.
We extract features from the \textbf{32 layers of Stage~3} as multi-layer conditioning.
The standard last-layer output (pre-RVQ) has dimensionality 768.

% ============================================================
\section{Method}
\label{sec:method}

\subsection{System Overview}

Our pipeline consists of three stages:

\textbf{(1) Offline feature extraction.}
All audio is processed once through DAC and MOSS before training:
\begin{itemize}[itemsep=2pt]
  \item DAC Encoder on clean speech $\to$ $\mathbf{x}_1 \in \mathbb{R}^{T \times 1024}$ (target)
  \item DAC Encoder on noisy speech $\to$ $\mathbf{x}_0 \in \mathbb{R}^{T \times 1024}$ (source)
  \item MOSS Encoder on noisy speech $\to$ conditioning features $\mathbf{c}$
\end{itemize}
This offline design ensures that training requires only loading pre-computed tensors---neither DAC (74M params) nor MOSS (1.6B params) are loaded during training, enabling Colab-scale compute.

\textbf{(2) Training.}
At each step, we sample $t \sim U(0,1)$, compute $\mathbf{x}_t$ (Eq.~\ref{eq:interpolation}), and train the DiT to predict the velocity field with the MSE loss (Eq.~\ref{eq:loss}).

\textbf{(3) Inference.}
Given a new noisy utterance, we encode it with both DAC (to get $\mathbf{x}_0$) and MOSS (to get $\mathbf{c}$), then run a 50-step Euler ODE from $t=0$ to $t=1$, and decode $\hat{\mathbf{x}}_1$ with the DAC decoder to produce the enhanced waveform.

\subsection{Diffusion Transformer Architecture}

The velocity predictor is a Diffusion Transformer (DiT):
\begin{itemize}[itemsep=2pt]
  \item \textbf{Input projection}: Linear($1024 \to 512$)
  \item \textbf{Timestep embedding}: Sinusoidal embedding $\to$ MLP($512 \to 2048 \to 512$)
  \item \textbf{6 DiT blocks}, each containing:
    \begin{enumerate}[itemsep=1pt]
      \item AdaLN($t$) $\to$ Multi-Head Self-Attention (8 heads)
      \item AdaLN($t$) $\to$ Cross-Attention (Q from latent, K/V from $\mathbf{c}$)
      \item AdaLN($t$) $\to$ Feed-Forward (4$\times$ expansion, GELU)
    \end{enumerate}
  \item \textbf{Output projection}: LayerNorm $\to$ Linear($512 \to 1024$)
\end{itemize}
Adaptive Layer Normalisation (adaLN) injects the timestep embedding as scale and shift parameters, following the standard DiT design~\cite{peebles2023scalable}.

\subsection{Multi-Layer Conditioning}
\label{sec:conditioning}

We compare four conditioning strategies that differ \emph{only} in how MOSS features are provided to the cross-attention layers.
The DiT backbone is otherwise identical across all variants.

\paragraph{Experiment 1: No conditioning (\texttt{none}).}
The cross-attention layers are removed entirely.
The DiT operates solely on the noisy DAC latent with timestep.
This serves as the baseline: 28.4M parameters.

\paragraph{Experiment 2: Last-layer conditioning (\texttt{last\_layer}).}
The standard approach: condition on $\mathbf{h}_{31} \in \mathbb{R}^{T_c \times 768}$, the final encoder output of MOSS before quantisation.
The cross-attention layers add +10.1M parameters (K/V projections from 768-dim condition to 512-dim hidden).

\paragraph{Experiment 3: Static multi-layer conditioning (\texttt{multi\_layer}).}
We extract all 32 hidden-layer outputs $\{\mathbf{h}_i\}_{i=0}^{31}$, each in $\mathbb{R}^{T_c \times 1280}$, and compute a learnable weighted sum:
\begin{equation}
  \mathbf{c} = \sum_{i=0}^{31} \text{softmax}(\mathbf{w})_i \cdot \mathbf{h}_i
  \label{eq:static-fusion}
\end{equation}
where $\mathbf{w} \in \mathbb{R}^{32}$ is a learnable parameter vector initialised to zeros (uniform weights after softmax).
This adds only 32 scalar parameters over the cross-attention baseline, but changes the conditioning dimension to 1280 (requiring adjusted K/V projections).

\paragraph{Experiment 4: Time-dependent multi-layer conditioning (\texttt{multi\_layer\_time}).}
Instead of static weights, a small MLP maps the ODE timestep to per-layer weights:
\begin{equation}
  \mathbf{c}(t) = \sum_{i=0}^{31} \text{softmax}\!\big(\text{MLP}(t)\big)_i \cdot \mathbf{h}_i
  \label{eq:time-fusion}
\end{equation}
The MLP has architecture $1 \xrightarrow{\text{Linear}} 64 \xrightarrow{\text{SiLU}} 32$, adding 2,208 parameters.
This allows the model to emphasise different types of features at different stages of the ODE trajectory---e.g., acoustic detail layers early in denoising and semantic layers later.

% ============================================================
\section{Experimental Setup}
\label{sec:setup}

\subsection{Data}

We use \textbf{LibriSpeech dev-clean} (2,703 utterances) as clean speech and \textbf{MUSAN noise/free-sound} as environmental noise.
Noisy mixtures are generated at SNR $\in \{-5, 0, 5, 10, 15\}$\,dB using our mixing pipeline.
As a proof-of-concept, all models are trained on the \textbf{5\,dB} condition.
The data is split 80/10/10 into train/validation/test sets (2,163/270/270 utterances).

All features are extracted offline:
\begin{itemize}[itemsep=2pt]
  \item DAC latents: $(T_{\text{dac}}, 1024)$ per utterance, truncated to 200 frames (4\,s at 50\,Hz)
  \item MOSS last-layer: $(T_{\text{moss}}, 768)$, MOSS multi-layer: 32 $\times$ $(T_{\text{moss}}, 1280)$
  \item MOSS features are truncated to 50 frames ($= 200/4$, matching the 50\,Hz/12.5\,Hz ratio)
\end{itemize}

\subsection{Training}

All four models are trained with identical hyperparameters:
\begin{itemize}[itemsep=2pt]
  \item AdamW optimiser, learning rate $10^{-4}$, weight decay $10^{-5}$
  \item Batch size 16, cosine annealing schedule with 200-step warmup
  \item Maximum 100 epochs, early stopping with patience 20
  \item Gradient clipping at norm 1.0
\end{itemize}
Training was performed on Google Colab Pro (T4/A100 GPUs).

\subsection{Evaluation Metrics}

We evaluate on three complementary metrics:

\textbf{PESQ} (Perceptual Evaluation of Speech Quality)~\cite{rix2001perceptual}: wideband mode, range $[-0.5, 4.5]$.
Approximates human mean opinion scores via an auditory model.
Note that generative flow-based models typically achieve lower PESQ than supervised discriminative models, as they optimise for naturalness rather than point-wise signal fidelity.

\textbf{STOI} (Short-Time Objective Intelligibility)~\cite{taal2010short}: range $[0, 1]$.
Computes the average correlation of short-time spectral envelopes between reference and degraded signals.

\textbf{FAD} (Fréchet Audio Distance)~\cite{kilgour2019frechet}: measures distributional similarity in VGGish embedding space:
\begin{equation}
  \text{FAD} = \|\boldsymbol{\mu}_r - \boldsymbol{\mu}_g\|^2 + \text{Tr}\!\left(\boldsymbol{\Sigma}_r + \boldsymbol{\Sigma}_g - 2(\boldsymbol{\Sigma}_r\boldsymbol{\Sigma}_g)^{1/2}\right)
\end{equation}
Lower FAD indicates more natural, reference-like audio.

% ============================================================
\section{Results}
\label{sec:results}

\subsection{Main Results}

Table~\ref{tab:results} presents the evaluation results across all four conditioning variants.

\begin{table}[h]
\centering
\caption{Evaluation results on the test set (270 utterances, 5\,dB SNR). Best results in \textbf{bold}.}
\label{tab:results}
\begin{tabular}{clccc}
\toprule
\textbf{Exp} & \textbf{Condition} & \textbf{PESQ $\uparrow$} & \textbf{STOI $\uparrow$} & \textbf{FAD $\downarrow$} \\
\midrule
1 & None (no MOSS)    & 1.6048 & 0.8527 & 2.9774 \\
2 & Last layer only   & 1.6499 \footnotesize{(+2.8\%)} & 0.8589 & 2.6997 \\
3 & Static multi-layer & 1.6868 \footnotesize{(+5.1\%)} & 0.8642 & 2.3857 \\
4 & Time-dep.\ multi-layer & \textbf{1.6986} \footnotesize{(+5.8\%)} & \textbf{0.8647} & \textbf{2.3456} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{All three metrics improve monotonically} from Experiment~1 through Experiment~4: no conditioning $<$ last layer $<$ static multi-layer $<$ time-dependent multi-layer.

\subsection{Analysis}

\paragraph{Multi-layer conditioning provides substantial gains.}
Moving from no conditioning (Exp~1) to last-layer conditioning (Exp~2) yields +2.8\% PESQ and $-9.3\%$ FAD.
Using all 32 layers (Exp~3) further improves to +5.1\% PESQ and $-19.9\%$ FAD, despite adding only 32 extra parameters over the cross-attention baseline.
This confirms our hypothesis that information distributed across multiple MOSS layers is complementary and beneficial.

\paragraph{Time-dependent fusion adds further improvement.}
Experiment~4 achieves the best results across all metrics.
The MLP produces time-varying weights at a cost of only 2,208 additional parameters.
Analysis of the learned time-dependent weights reveals:
\begin{itemize}[itemsep=2pt]
  \item At $t \approx 0$ (start of ODE, close to the noisy input), weights are approximately uniform across all 32 layers.
  \item As $t \to 1$ (approaching clean target), the weights become more peaked, with increased emphasis on earlier layers.
  \item The standard deviation of the weight distribution increases from 0.0012 at $t=0$ to 0.0031 at $t=1$, indicating progressive specialisation.
\end{itemize}
This behaviour is consistent with an intuitive interpretation: early in the denoising trajectory, the model benefits from a broad mixture of information, while later stages require more targeted acoustic detail (captured by lower layers).

\paragraph{Static fusion learns near-uniform weights.}
In Experiment~3, the learned weight vector has a Gini coefficient of only 0.007 (near-perfectly uniform).
All 32 layers contribute roughly equally, suggesting that each layer adds non-redundant information.
This uniformity also motivates the time-dependent variant: if all layers matter, giving the model the flexibility to dynamically re-weight them per timestep can only help.

\paragraph{PESQ score context.}
Our PESQ scores (1.60--1.70) may appear modest compared to supervised discriminative models.
However, flow-based generative models optimise for perceptual naturalness rather than point-wise signal reconstruction.
The FAD metric, which measures distributional quality, shows a much larger improvement ($-21.2\%$ from Exp~1 to Exp~4), indicating that multi-layer conditioning most significantly improves the perceptual naturalness and distributional fidelity of the enhanced speech.

% ============================================================
\section{Limitations and Future Work}
\label{sec:limitations}

\subsection{Limitations}

\paragraph{Single SNR training.}
All models were trained on 5\,dB SNR only, as a proof-of-concept under constrained compute.
While noisy mixtures at $\{-5, 0, 5, 10, 15\}$\,dB are available, a single-SNR model may not generalise across the full noise-level spectrum.

\paragraph{Limited noise diversity.}
The MUSAN free-sound subset provides environmental noise but does not cover the full diversity of real-world acoustic conditions (e.g., reverberation, multi-speaker babble, music interference).
Models trained on this data will likely struggle with out-of-distribution noise types.

\paragraph{Computational cost of multi-layer conditioning.}
Using all 32 MOSS layers means $32\times$ more conditioning data per utterance compared to single-layer conditioning.
During feature extraction, this increases storage (32 tensors of $(T, 1280)$ vs.\ one of $(T, 768)$); at training time, the fusion operation adds modest compute.

\paragraph{DAC reconstruction ceiling.}
Since we operate in DAC's continuous latent space, the maximum achievable quality is bounded by DAC's own reconstruction fidelity.
Any artefacts introduced by DAC encoding and decoding will appear in our enhanced output regardless of the flow model's quality.

\subsection{Future Work}

\begin{enumerate}[itemsep=3pt]
  \item \textbf{Multi-SNR training} using the available $\{-5, 0, 5, 10, 15\}$\,dB data for a robust, general-purpose model.
  \item \textbf{Layer selection study}: the near-uniform weights in Exp~3 suggest all layers contribute, but a subset (e.g., first~15 + last) may achieve comparable results at lower cost.
  \item \textbf{Scaling}: larger datasets (full LibriSpeech, DNS Challenge) and more noise types would test generalisation.
  \item \textbf{Faster inference} via ODE distillation (e.g., consistency distillation~\cite{song2023consistency}) to reduce the 50-step Euler solver.
  \item \textbf{Alternative fusion architectures}: attention-based layer fusion or per-block layer selection in the DiT.
\end{enumerate}

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have presented a flow-matching speech enhancement system that operates in the continuous latent space of pre-trained audio tokenizers (DAC and MOSS).
Through a controlled four-way ablation, we demonstrate that \emph{multi-layer} conditioning from the MOSS tokenizer consistently outperforms both unconditional and last-layer-only baselines across all metrics.
Our time-dependent fusion variant, which dynamically adjusts per-layer weights based on the ODE timestep, achieves the best results at a negligible parameter cost (+2,208 parameters).

The monotonic improvement pattern---no conditioning $<$ last layer $<$ static multi-layer $<$ time-dependent multi-layer---provides strong evidence that leveraging the full depth of pre-trained audio encoders is beneficial for speech enhancement, analogous to multi-scale feature extraction in computer vision.

% ============================================================
\bibliographystyle{plainnat}
\begin{thebibliography}{10}

\bibitem[Lipman et~al.(2023)]{lipman2023flow}
Yaron Lipman, Ricky T.~Q. Chen, Heli Ben-Hamu, Maximilian Nickel, and Matt Le.
\newblock Flow matching for generative modeling.
\newblock In \emph{International Conference on Learning Representations (ICLR)}, 2023.

\bibitem[Liu et~al.(2023)]{liu2023flow}
Xingchao Liu, Chengyue Gong, and Qiang Liu.
\newblock Flow straight and fast: Learning to generate and transfer data with rectified flow.
\newblock In \emph{International Conference on Learning Representations (ICLR)}, 2023.

\bibitem[Kumar et~al.(2024)]{kumar2024high}
Rithesh Kumar, Prem Seetharaman, Alejandro Luebs, Ishaan Kumar, and Kundan Kumar.
\newblock High-fidelity audio compression with improved {RVQGAN}.
\newblock In \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2023.

\bibitem[Ma et~al.(2024)]{ma2024language}
Ziyang Ma, Yakun Song, Chenpeng Du, Jian Cong, Zhuo Chen, Yuping Wang, Yuxuan Wang, and Xie Chen.
\newblock Language model can listen while speaking.
\newblock \emph{arXiv preprint arXiv:2408.02622}, 2024.

\bibitem[Peebles and Xie(2023)]{peebles2023scalable}
William Peebles and Saining Xie.
\newblock Scalable diffusion models with transformers.
\newblock In \emph{International Conference on Computer Vision (ICCV)}, 2023.

\bibitem[Richter et~al.(2023)]{richter2023speech}
Julius Richter, Simon Welker, Jean-Marie Lemercier, Bunlong Lay, and Timo Gerkmann.
\newblock Speech enhancement and dereverberation with diffusion-based generative models.
\newblock \emph{IEEE/ACM Transactions on Audio, Speech, and Language Processing}, 2023.

\bibitem[Lu et~al.(2013)]{lu2013speech}
Xugang Lu, Yu Tsao, Shigeki Matsuda, and Chiori Hori.
\newblock Speech enhancement based on deep denoising autoencoder.
\newblock In \emph{Interspeech}, 2013.

\bibitem[Rix et~al.(2001)]{rix2001perceptual}
Antony~W Rix, John~G Beerends, Michael~P Hollier, and Andries~P Hekstra.
\newblock Perceptual evaluation of speech quality ({PESQ}) --- a new method for speech quality assessment of telephone networks and codecs.
\newblock In \emph{IEEE International Conference on Acoustics, Speech, and Signal Processing (ICASSP)}, 2001.

\bibitem[Taal et~al.(2010)]{taal2010short}
Cees~H Taal, Richard~C Hendriks, Richard Heusdens, and Jesper Jensen.
\newblock A short-time objective intelligibility measure for time-frequency weighted noisy speech.
\newblock In \emph{IEEE International Conference on Acoustics, Speech, and Signal Processing (ICASSP)}, 2010.

\bibitem[Kilgour et~al.(2019)]{kilgour2019frechet}
Kevin Kilgour, Mauricio Zuluaga, Dominik Roblek, and Matthew Sharifi.
\newblock Fr\'echet audio distance: A reference-free metric for evaluating music enhancement algorithms.
\newblock In \emph{Interspeech}, 2019.

\bibitem[Song et~al.(2023)]{song2023consistency}
Yang Song, Prafulla Dhariwal, Mark Chen, and Ilya Sutskever.
\newblock Consistency models.
\newblock In \emph{International Conference on Machine Learning (ICML)}, 2023.

\end{thebibliography}

\end{document}
